Investigating country-level activity and mobility patterns and their interdependencies on COVID-19 outcomes

### 2020-09-30 docs/manuscript/manuscript.tex
comments on manuscript added

\usepackage{graphicx,booktabs,calc}

%=== GRAPHICS PATH ===========
%\graphicspath{{./140408-Images/}}
% Marginpar width
%Marginpar width
\newcommand{\pts}[1]{\marginpar{ \small\hspace{0pt} \textit{[#1]} } }
\setlength{\marginparwidth}{.5in}
%\reversemarginpar
%\setlength{\marginparsep}{.02in}

@@

\newcommand{\osn}{\oldstylenums}
\newcommand{\dg}{^{\circ}}
\newcommand{\lt}{\left}
\newcommand{\rt}{\right}
\newcommand{\pt}{\phantom}
\newcommand{\tf}{\therefore}
\newcommand{\?}{\stackrel{?}{=}}
\newcommand{\fr}{\frac}
\newcommand{\dfr}{\dfrac}
\newcommand{\ul}{\underline}
\newcommand{\tn}{\tabularnewline}
\newcommand{\nl}{\newline}
\newcommand\relph[1]{\mathrel{\phantom{#1}}}
\newcommand{\cm}{\checkmark}
\newcommand{\ol}{\overline}

%%% COLORS %%%%
\newcommand{\rd}{\color{red}}
\newcommand{\bl}{\color{blue}}
\newcommand{\pl}{\color{purple}}
\newcommand{\og}{\color{orange!90!black}}
\newcommand{\gr}{\color{green!40!black}}
% Custom colors
\usepackage{color}
\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}

@@

\section{Introduction}
COVID-19 has had a severe impact on mobility and social activities since initial reports of the disease emerged in late 2019.
By May 2020 most countries had implemented social distancing measures and imposed strict travel restrictions to curtail the spread of the virus
both within their borders and internationally.
According to \cite{bonaccorsi2020economic}, the international spread of the disease can be partially explained by risk perception,
mobility behavior and social media influences using a global vector autoregression (GVAR) modeling framework\cite{dees2007exploring}.
%Milani concludes that most countries learn from each other's social distancing response which can be partially explained by social networks.  
\cite{zhang2020pathways} confirm that human mobility is key to understanding the transmission pathways.
Using a mobility dataset of over 500,000 flights and over 100 million passengers,
the authors develop a dynamic network model using time-dependent border restrictions as exogenous variables.
\cite{mckenzie2020country} studied the patterns in 6 activity responses across 108 countries but this was an exploratory effort in comparison with government policies and it included no modeling attempt to explain COVID-19 outcomes.

@@

\section{Literature review}

@@

\subsection{Dynamic time warping}
The dynamic time warping (DTW) algorithm \cite{giorgino2009computing} computes the optimal distance between pairwise time series.
Any hierarchical clustering procedure requires a dissimilarity matrix as an input, which DTW provides.
First, we compute the Euclidean distance matrix for each country across the dimensions of the endogenous variables.
Second, we apply DTW to find the optimal matching between the countries based on their pairwise distance matrices across the dimensions observed.
The symmetric dissimilarity matrix is then used an input to the clustering algorithm.

\subsection{Clustering}
We use the Ward method \cite{murtagh2014ward}, which is a hierarchical agglomerative clustering (HAC) approach, to group the countries.
The dendogram is shown in \autoref{fig:dend}.

@@

\subsection{Vector autoregression model}
The basic vector autoregression (VAR) model is given by:
\begin{equation}
  \label{eq:6}
  y_{t} = A_{0}(t) + A(\ell)Y_{t-1} + \varepsilon_{t}
\end{equation}
where $y_{t}$ is a $k\times 1$ vector of endogenous variables for a given unit at time $t\in \{1,\ldots, T\}$,
$A_{0}(t)$ captures the deterministic components and 
$A(\ell)$ is a polynomial in the lag operator and $\varepsilon_{t}\sim iid(0,\Sigma_{\varepsilon})$ are random disturbances.
This framework allows for specification of both dynamic and static interdependencies
among the endogenous variables, as well as cross-sectional heterogeneities (when a panel VAR framework is used).
Country-specific VAR models can also be stacked to yield a global VAR  \cite{martin2017weighting,dees2007exploring,covid19b}, allowing for the inclusion of weakly exogenous (``foreign'') variables (such as border closures) using weights (such as air travel flows) from each of the countries.

@@

\section{Conclusion and future work}
We cluster the countries in our sample into four categories based on similar activity and mobility trends.
% By running the model with a different lag parameter for each country we attempted to estimate the optimal lag for each country.
% However, our VAR implementation fell short of producing statistically significant results.
Future work will focus on estimating VAR models for each of the countries, adding exogenous and weakly exogenous variables such as interventions and restrictions (social distancing orders, travel bans, etc.), as well as deterministic ones (country characteristics).
Subsequently, cluster-level VARs and a global VAR will be estimated.
These models will provide insights into the processes governing the changes in the endogenous variables (COVID-19 outcomes, activity and mobility trends), as well as potential forecasting capabilities.
The insights gained from these models would enable policy and decision makers plan efficiently for recovery from the current pandemic, as well as better prepare for future ones, by tailoring interventions relevant to the behavioral profiles of their respective countries.
%We also plan to substitute the number of new cases with a moving average for new COVID-19 cases which would produce more robust results.

>>>>

\usepackage{booktabs,calc,soul,xcolor}

% Marginpar width
%Marginpar width
\newcommand{\pts}[1]{\marginpar{ \small\hspace{0pt} \textit{[#1]} } }
\setlength{\marginparwidth}{.5in}
%\reversemarginpar
%\setlength{\marginparsep}{.02in}

@@

\newcommand{\osn}{\oldstylenums}
\newcommand{\dg}{^{\circ}}
\newcommand{\lt}{\left}
\newcommand{\rt}{\right}
\newcommand{\pt}{\phantom}
\newcommand{\tf}{\therefore}
\newcommand{\?}{\stackrel{?}{=}}
\newcommand{\fr}{\frac}
\newcommand{\dfr}{\dfrac}
\newcommand{\tn}{\tabularnewline}
\newcommand{\nl}{\newline}
\newcommand\relph[1]{\mathrel{\phantom{#1}}}
\newcommand{\cm}{\checkmark}
\newcommand{\ol}{\overline}

%%% COLORS %%%%
\newcommand{\rd}{\color{red}}
\newcommand{\bl}{\color{blue}}
\newcommand{\pl}{\color{purple}}
\newcommand{\og}{\color{orange!90!black}}
\newcommand{\gr}{\color{green!40!black}}
% Custom colors
%\usepackage{color}
\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}

@@

\section{Introduction}
\textit{\bl This section sets up the motivation and significance of the work presented in this paper. Key questions:
  \begin{enumerate}
  \item What is the ongoing and potential future impact of COVID-19 across various countries, particularly on mobility and activity patterns? What are the implications (known and unknown)?
  \item Briefly what has been done in terms of modeling mobility impacts?
  \item Where are the current gaps in our knowledge?
  \item (Final paragraph) What is the summary of this paper's contributions? What is the expected significance of our results?
  \end{enumerate}
}

COVID-19 has had a severe impact on mobility and social activities since initial reports of the disease emerged in late 2019.
By May 2020 most countries had implemented social distancing measures and imposed strict travel restrictions to curtail the spread of the virus
both within their borders and internationally.
According to \cite{bonaccorsi2020economic}, the international spread of the disease can be partially explained by risk perception,
mobility behavior and social media influences using a global vector autoregression (GVAR) modeling framework\cite{dees2007exploring}.
%Milani concludes that most countries learn from each other's social distancing response which can be partially explained by social networks.  
\cite{zhang2020pathways} confirm that human mobility is key to understanding the transmission pathways.
Using a mobility dataset of over 500,000 flights and over 100 million passengers,
the authors develop a dynamic network model using time-dependent border restrictions as exogenous variables.
\cite{mckenzie2020country} studied the patterns in 6 activity responses across 108 countries but this was an exploratory effort in comparison with government policies and it included no modeling attempt to explain COVID-19 outcomes.

@@

\subsection{Related work}
\hl{Here, 3-4 paragraphs summarizing efforts to analyze activity-mobility patterns (i) historically, (ii) in context of COVID-19, and then (iii) brief history and application of VAR modeling.}

@@

\subsection{Dynamic time warping and clustering}
\hl{Not sure if we need to perform any exploratory analysis in this manuscript, but these methods might be useful either initially or in analyzing model results}
The dynamic time warping (DTW) algorithm \cite{giorgino2009computing} computes the optimal distance between pairwise time series.
Any hierarchical clustering procedure requires a dissimilarity matrix as an input, which DTW provides.
First, we compute the Euclidean distance matrix for each country across the dimensions of the endogenous variables.
Second, we apply DTW to find the optimal matching between the countries based on their pairwise distance matrices across the dimensions observed.
The symmetric dissimilarity matrix is then used an input to the clustering algorithm.

We use the Ward method \cite{murtagh2014ward}, which is a hierarchical agglomerative clustering (HAC) approach, to group the countries.
The dendogram is shown in \autoref{fig:dend}.

\subsection{Global vector autoregression model with time-varying weights}

@@

The basic vector autoregression (VAR) model is given by:
\begin{equation}
  \label{eq:6}
  y_{t} = A_{0}(t) + A(\ell)Y_{t-1} + \varepsilon_{t}
\end{equation}
where $y_{t}$ is a $k\times 1$ vector of endogenous variables for a given unit at time $t\in \{1,\ldots, T\}$,
$A_{0}(t)$ captures the deterministic components and 
$A(\ell)$ is a polynomial in the lag operator and $\varepsilon_{t}\sim iid(0,\Sigma_{\varepsilon})$ are random disturbances.
This framework allows for specification of both dynamic and static interdependencies
among the endogenous variables, as well as cross-sectional heterogeneities (when a panel VAR framework is used).
Country-specific VAR models can also be stacked to yield a global VAR  \cite{martin2017weighting,dees2007exploring,covid19b}, allowing for the inclusion of weakly exogenous (``foreign'') variables (such as border closures) using weights (such as air travel flows) from each of the countries.

@@

\section{Conclusion }
We cluster the countries in our sample into four categories based on similar activity and mobility trends.
% By running the model with a different lag parameter for each country we attempted to estimate the optimal lag for each country.
% However, our VAR implementation fell short of producing statistically significant results.
Future work will focus on estimating VAR models for each of the countries, adding exogenous and weakly exogenous variables such as interventions and restrictions (social distancing orders, travel bans, etc.), as well as deterministic ones (country characteristics).
Subsequently, cluster-level VARs and a global VAR will be estimated.
These models will provide insights into the processes governing the changes in the endogenous variables (COVID-19 outcomes, activity and mobility trends), as well as potential forecasting capabilities.
The insights gained from these models would enable policy and decision makers plan efficiently for recovery from the current pandemic, as well as better prepare for future ones, by tailoring interventions relevant to the behavioral profiles of their respective countries.
%We also plan to substitute the number of new cases with a moving average for new COVID-19 cases which would produce more robust results.
